---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Nowcasting examples

This repository contains two example approaches to nowcasting, one using the [EpiNow2](https://epiforecasts.io/EpiNow2/) package, and one using the [epinowcast](https://epiforecasts.io/epinowcast/) package which is currently being developed to eventually replace `EpiNow2` for real-time applications.

## Installation

You can install the utility functions contained in this repo with

```{r install, eval = FALSE}
# install.packages("devtools")
devtools::install_github("epiforecasts/nowcasting.example")
```

## Render the `EpiNow2` report

To render the report, run

``` r
rmarkdown::render("inst/reports/epinow2.Rmd")
```

Afterwards the report can be viewed in
`inst/reports/epinow2.html`. A [rendered version](inst/reports/epinow2.md) of the report is also available in the repository.

## Render the `epinowcast` report

To render the report, run

``` r
rmarkdown::render("inst/reports/epinowcast.Rmd")
```

Afterwards the report can be viewed in
`inst/reports/epinowcast.html`. A [rendered version](inst/reports/epinowcast.md) of the report is also available in the repository.

## Create mock dataset

A mock dataset can be created using

```{r data, eval = FALSE}
source("inst/scripts/create_mock_dataset.r")
```


